---
title: "A Complete Survival Analysis Case Study: the Lung Data set"
author: "Sophie Gallet"
output: html_notebook
---

# Case Study: Lung Data set

## Data Preparation
```{r}
library(tidyverse)
library(survival)
?lung
```

```{r}
table(lung$inst)
```

```{r}
nrow(lung)
```

```{r}
d_raw <- as_tibble(lung)
summary(d_raw)
```

```{r}
d<-mutate(d_raw,
          event = 0 + (status==2), #change values from 0/2 to 0/1 - censoring
          inst = factor(inst), #change type
          sex = factor(sex, levels = 1:2, labels=c("male", "female")) #readable levels
          )
summary(d)
```
### Deal with missing values

```{r}
# 47 missing for meal.cal, options: remove feature, impute, (remove observations)
fit.meal <- lm(meal.cal ~ sex, data = d) #replace missing values with per sex average
d$meal.cal[is.na(d$meal.cal)] <-
  predict(fit.meal, newdata = subset(d, is.na(meal.cal)))
summary(d)
```

```{r}
d$wt.loss[is.na(d$wt.loss)] <- 
  predict(lm(wt.loss ~ age + sex, data = d), newdata = subset(d, is.na(wt.loss)))

summary(d)
```

One observation has no institution - if it was our study we'd want to find out which institution that was. 
```{r}
#d[is.na(d$inst),] same thing as:
subset(d, is.na(d$inst))
```

```{r}
d$y <-with(d, Surv(time/30.5, event)) #number of months 
head(d$y)
```

## Exploratory Data Analysis

```{r}
survfit(y~1, data=d)
```

```{r}
survfit(y ~ sex, data = d)
```

```{r}
str(d)
```

```{r}
table(d$ph.ecog) #0:good, 5:dead
```
```{r}
survfit(y ~ ph.ecog, data = d)
```

As expected a higher score leads to a lower survival time. 

```{r}
hist(d$age)
```

```{r}
d$ageCat <- cut(d$age, breaks = c(0,50,70,Inf))
table(d$ageCat)
survfit(y ~ ageCat, data = d)
```

No big difference between age groups. 

```{r}
table(d$ph.karno)
table(d$pat.karno)
survfit(y ~ ph.karno, data = d)
survfit(y ~ pat.karno, data = d)
```

Some of the medians are not too reliable because we have small numbers in the groups.

```{r}
hist(d$meal.cal)
```
```{r}
survfit(y ~ I(meal.cal < 800), data = d)
```
No big difference in survival between the two groups, the confidence intervals pretty overlap to a large degree. 

```{r}
stem(d$wt.loss)
survfit(y ~ I(wt.loss > 0), data = d)
stem(d$ph.karno)
stem(d$meal.cal)
survfit(y ~ I(meal.cal < 800), data = d)
stem(d$wt.loss)
```

### Comparing 2 groups

```{r}
survdiff(y ~ I(wt.loss > 0), data = d)
survdiff(y ~ ageCat, data = d)
survdiff(y ~ sex, data = d)
survdiff(y ~ I(meal.cal < 800), data = d)
survdiff(y ~ ph.ecog, data = d) #treating them as categorical variables
survdiff(y ~ ph.karno, data = d)
```

Let's organize our logrank test results in a more compact manner:

```{r}
test_variable <- function(var_name) {
  e$x <- e[[ var_name ]]
  survdiff(y ~ x, data = e)
}

e <- mutate(d,
  weight_loss = wt.loss > 0,
  meal_calories_low = meal.cal < 800,
  age = ageCat
)

logrank_tests <-
  tibble(variable = c("weight_loss", "age", "sex", "meal_calories_low", "ph.ecog", "ph.karno")) %>%
  mutate(obj = map(variable, test_variable),
         tab = map(obj, broom::glance)) %>%
  unnest(tab)
logrank_tests
```

## Data Modeling and Machine Learning
### Model training
```{r}
load("data/lung.RData")
d1 <- lung
fit <- coxph(y ~ ., data = d1)
summary(fit)
```

Interpretation of coef: for 1 unit increase of ph.karno while keeping other variables fixed, we get coef_value increase in survival.

Note that the exp(coef) can help us interpret the data (with everything else the same, on average a male patient's is twice as likely to die (?) and inversely a female is 50% less likely to die). 

### model 1. AIC-STEP
```{r}
fit.aic <- step(fit)
summary(fit.aic)

b.aic <- coef(fit.aic) #saving coef for future comparison
```

### model 2. manual
Based on expertise on the field, we (pretend to) use expert opinion to build a model:
```{r}
fit.manual <- coxph(y ~ sex + ph.ecog + pat.karno + wt.loss, data = d1)
summary(fit.manual)

b.manual <- coef(fit.manual) #saving the coef for future use
```
### model 3.elastic net
```{r}
library(glmnet)
X<-model.matrix(y ~ ., data = d1)[,-1]
str(X)
head(X)
y<-d1$y

```

```{r}
fit <- cv.glmnet(X, y, family = "cox")
b.enet.all <- coef(fit, s="lambda.min")
b.enet <- b.enet.all[b.enet.all != 0] #keep only non zero coeff
names(b.enet) <- colnames(X)[as.logical(b.enet.all != 0)]
b.enet
```

### model 4. CCP

```{r}
#adply: for each slice of an array, apply function then combine results into a data frame. 2: split by cols
#Tidy the result of a test into a summary data.frame, with disposable row names: suited for manip by plyr
fits <- plyr::adply(X, 2, function(x) broom::tidy(coxph(y ~ x))) 
print(fits)
#str(fits)

b.CCP <- with(fits, structure(estimate, names = as.character(X1)))
b.CCP
```

```{r}
# Combine all coeffs together 
models_coeff <- tibble(
  method = c("manual", "aic", "elnet", "cpp"),
  coefficients = list(b.manual, b.aic, b.enet, b.CCP)
)
models_coeff
```


## Model testing

```{r}
# sweep returns an array obtained from an input array(x) by sweeping out a summary statistic (b)
lincom <- function(b,X) rowSums(sweep(x = X[ , names(b), drop=FALSE], 
                                      MARGIN = 2, 
                                      STATS = b, 
                                      FUN="*"))
```
```{r}
load("data/lung_newdata.RData")
X.new <- model.matrix(y ~ . - 1, lung_newdata)
y <- lung_newdata$y
```

```{r}
models_performance <- mutate(models_coeff,
  predictions = map(coefficients, ~ lincom(., X.new)),
  cox_obj = map(predictions, ~ coxph(y ~ I(. / sd(.)))),
  cox_tab = map(cox_obj, broom::tidy)
) %>%
  unnest(cox_tab)
models_performance
```
## Sharing the results outside of R
```{r}
models_coefficients_flat <- mutate(models_coeff,
  coefficients_tab = map(coefficients, ~ tibble(feature = names(.), 
                                                coefficient = unname(.)))) %>%
  unnest(coefficients_tab, .drop = TRUE) %>%
  select(method, feature, coefficient)
models_coefficients_flat
```

Write tables on disk:
```{r}
write_csv(models_coefficients_flat, "models_coefficients.csv")
write_csv(models_performance, "models_performance.csv")
```

