---
title: "Survival_analysis_theory"
author: "Sophie Gallet"
date: "10/22/2019"
output: html_document
---

#Time to Event Analysis / Survival Analysis in R

Based on datacamp class. 

## Intro: What is Survival Analysis 

We're interested in the time until an event occurs (ex: time taxi takes to arrive after being called)

Why do we need special methods for SA?
- duration times are always positive. We need distributions that can work with that
- different measures of interest
- **censoring** - when we don't have the time event occured (ex: patient dropped out of clinical study)

```{r}
library("survival")
library("survminer") #for more advanced visualization
```

```{r}
# Check out the help page for this dataset
#help(GBSG2, package = "TH.data")

# Load the data
data(GBSG2, package = "TH.data")

# Look at the summary of the dataset
summary(GBSG2)
```

### Censoring

```{r}
# Testing surv()
time <-c(5,6,2,4,4)
event <-c(1,0,0,1,1) #0 if censored data


Surv(time, event) 
```

```{r}
# Count censored and uncensored data
num_cens <- table(GBSG2$cens)
num_cens

# Create barplot of censored and uncensored data
barplot(num_cens)

# Use help() to look at cens
help(GBSG2, package = "TH.data")
```

```{r}
# Create Surv-Object
sobj <- Surv(GBSG2$time, GBSG2$cens)

# Look at 10 first elements
sobj[1:10]

# Look at summary
summary(sobj)

# Look at structure
str(sobj)
```

```{r}
# Load the UnempDur data
data(UnempDur, package = "Ecdat")

# Count censored and uncensored data
cens_employ_ft <- table(UnempDur$censor1)
cens_employ_ft

# Create barplot of censored and uncensored data
barplot(cens_employ_ft)

# Create Surv-Object
sobj <- Surv(UnempDur$spell, UnempDur$censor1) #Surv() to define time to event outcomes

# Look at 10 first elements
sobj[1:10]
```

### Measures used in SA

We're mostly interested in the survival function or survival curve.

Allows us to answer questions such as:
- What is the proba that a breast cancer patient survives more than 5 years?
- What's the typical waiting tiem for a taxi in NYC?

$S(t) = 1-F(t) = P(T>t)$ 

Interpretation: probability that the duration is longer than t. 
Can also be used to find median time (for which S(t)=.5)

## Estimation of survival curves

### Kaplan-Meier Estimates

$\hat{S}(t) = \prod_{i:t_i \le t} \frac{n_i - d_i}{n_i} $

where $n_i$ is the number of individuals under observations at time $t_i$
and $d_i$ is the number of individuals who die (in our context) at time $t_i$

```{r}
# Using test data
time <-c(5,6,2,4,4)
event <-c(1,0,0,1,1) #0 if censored data

# Manually computation of Shat
shat2 = (5-0)/5 
shat3 = shat2 * (4-0)/4 
shat4 = shat3 * (4-2)/4
shat5 = shat4 * (2-1)/2
shat6 = shat5 * (1-0)/1


# Coding the function - helps me understand/remember it
shat <- function(t, times, events){
  shat = 1 #meaninful initial value
  ni = length(times) # we start with all the observations 
  for (ti in 1:t) {
    di = sum(events[times==ti]) #number of individuals dying (in our case) at ti
    ci = sum(events[times==ti]==0) #censored events at ti
    shat = shat * (ni-di)/ni
    ni = ni-di-ci #number of individuals remaining at ti+1 
  }
  return(shat)
}
print("for t=4")
print(c("Manual computation: " = shat4,
        "Function computation: " = shat(4, time, event)))
```
```{r}
# The R function to estimate survival curves 
km <-survfit(Surv(time, event)~1) 
summary(km)
# plot function from survminer package - bug, missing data argument (see below)
#ggsurvplot(km, conf.int=FALSE, risk.table="nrisk_cumevents", legend="none")
```
```{r}
# Create time and event data
time <- c(5, 6, 2, 4, 4)
event <- c(1, 0, 0, 1, 1)

# Compute Kaplan-Meier estimate
km <- survfit(Surv(time, event) ~ 1)
km

# Take a look at the structure
str(km)

# Create data.frame
data.frame(time = km$time, n.risk = km$n.risk, n.event = km$n.event, 
n.censor = km$n.censor, surv = km$surv)
```

### Understanding and Visualizing Kaplan-Meier curves

Ex: You throw a party and at 1 a.m. guests suddenly start dancing. You are curious to analyze how long your guests will dance for and start collecting data. The problem is that you get tired and go to bed after a while. You obtain the following right censored dancing times data given in dancedat:
- name is the name of you r friend.
- time is the right-censored dancing time.
- obs_end indicates if you observed the end of your friends dance (1) or if you went to sleep before they stopped dancing (0).

```{r}
# Create dancedat data
dancedat <- data.frame(
  name = c("Chris", "Martin", "Conny", "Desi", "Reni", "Phil", 
    "Flo", "Andrea", "Isaac", "Dayra", "Caspar"),
  time = c(20, 2, 14, 22, 3, 7, 4, 15, 25, 17, 12),
  obs_end = c(1, 1, 0, 1, 1, 1, 1, 1, 0, 0, 0))

# Estimate the survivor function pretending that all censored observations are actual observations.
km_wrong <- survfit(Surv(time) ~ 1, data = dancedat)

# Estimate the survivor function from this dataset via kaplan-meier.
km <- survfit(Surv(time, obs_end) ~ 1, data = dancedat)

# Plot the two and compare
ggsurvplot_combine(list(correct = km, wrong = km_wrong))
```

```{r}
# Kaplan-Meier estimate
km <- survfit(Surv(time, cens)~1, data = GBSG2)

# plot of the Kaplan-Meier estimate
ggsurvplot(km)

# add the risk table to plot
ggsurvplot(km, risk.table = TRUE)

# add a line showing the median survival time
ggsurvplot(km, risk.table = TRUE, surv.median.line = "hv")
```

### The Weibull model

Kaplan-Meier is powerful, used by many people, but mostly as a descriptive tool to look at the data. It approximates the survival curve with a step function, while the Weibull model creates a smooth function. 
```{r}
# Weibull model
wb <-survreg(Surv(time, cens)~1, data =GBSG2)

# 90% of patients survive beyong time point:
predict(wb, type ="quantile", p = 1-.9, newdata = data.frame(1)) #1-0.9 bc F(t)=1-S(t)

# Compute the median survival from the model
#predict(wb, type = "quantile", p = .5, newdata = data.frame(1))
```
90% of patients survive more than 384.99 days.

```{r}
# Weibull model
wb <- survreg(Surv(time, cens) ~ 1, data=GBSG2)

# Retrieve survival curve from model probabilities 
surv <- seq(.99, .01, by = -.01)

# Get time for each probability
t <- predict(wb, type = "quantile", p = 1-surv, newdata = data.frame(1))

# Create data frame with the information
surv_wb <- data.frame(time = t, surv = surv)

# Look at first few lines of the result
head(surv_wb)
```

#### Visualizing the Weibull model

Visualization functions often focus on step functions. 
```{r}
# Weibull model
wb <- survreg(Surv(time, cens)~1, data=GBSG2)

# Retrieve survival curve from model
surv <- seq(.99, .01, by = -.01)

# Get time for each probability
t <- predict(wb, type = "quantile", p = 1-surv, newdata = data.frame(1))

# Create data frame with the information needed for ggsurvplot_df
surv_wb <- data.frame(time = t, surv = surv, 
  upper = NA, lower = NA, std.err = NA)

# Plot
ggsurvplot_df(fit = surv_wb, surv.geom = geom_line)
```
50% of patients survive more than 1693.93 days

## The Weibull model

### Why Weibull

Look at various survival curves 

```{r}
wbmod <- survreg(Surv(time, cens) ~ horTh + tsize, data = GBSG2) # we replace  1 by the exp variables
coef(wbmod) # access coeff
```

A positive effect suggests an increased time until the event (in our case, death happens). Receiving hormone therapy leads to prolonged survival, while a bigger tumor size leads to a decreased survival. 

### Visualizing Weibull models

Steps to produce visualization
- compute weibull model
- decide on "imaginary patients (combination of tsize and horTh)
- compute survival curves for imaginary patients
- create data.frame with survival curve information
- plot

```{r}
# Weibull model
wbmod <- survreg(Surv(time, cens) ~ horTh + tsize, data = GBSG2)

# Imaginary patients
newdat <- expand.grid(
  horTh = levels(GBSG2$horTh),
  tsize = quantile(GBSG2$tsize, probs = c(.25, .5, .75)))

# Compute survival curves
surv <- seq(.99, .01, by=-.01)
t<- predict(wbmod, type="quantile", p=1-surv, newdata = newdat)

# How many rows and columns does t have?
dim(t) 
```

Each row of t corresponds to one covariate combination (one imaginary patient) and each column to one value of surv. 

```{r}
library(reshape2)
# Use cbind() to combine the information in newdat with t
surv_wbmod_wide <- cbind(newdat, t)
  
# Use melt() to bring the data.frame to long format
surv_wbmod <- melt(surv_wbmod_wide, id.vars = c("horTh", "tsize"), 
                   variable.name = "surv_id", value.name = "time")

# Use surv_wbmod$surv_id to add the correct survival probabilities surv
surv_wbmod$surv <- surv[as.numeric(surv_wbmod$surv_id)]

# Add columns upper, lower, std.err, and strata to the data.frame
surv_wbmod[, c("upper", "lower", "std.err", "strata")] <- NA

# Plot the survival curves
ggsurvplot_df(surv_wbmod, surv.geom = geom_line,
  linetype = "horTh", color = "tsize", legend.title = NULL)



```

### Other distributions than Weibull

Options include exponential, lognormal. The exponential model is a special case of the Weibull, and Weibull is more flexible. 

```{r}
# Weibull model
wbmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2)

# Lognormal model
lnmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2, dist="lognormal")

# newdata
newdat<-data.frame(horTh = levels(GBSG2$horTh))

# survival probabilities
surv <- seq(.99, .01, by = -.01)

# Survival curve from Weibull model and log-normal model
wbt <- predict(wbmod, type ="quantile", p=1-surv, newdata = newdat)
lnt <- predict(lnmod, type ="quantile", p=1-surv, newdata = newdat)

# Combine the information in newdat with wbt, lnt
surv_wb<- cbind(newdat, wbt, "dist" = "weibull")
surv_ln<- cbind(newdat, lnt, "dist" = "lognormal")

surv_wide <- rbind(surv_wb, surv_ln)
dim(surv_wide) # 4 rows, the number of covariate combinations 
```


```{r}
# Melt the data.frame into long format.
surv_long <- melt(surv_wide, id.vars = c("horTh", "dist"), variable.name = "surv_id", value.name = "time")

# Add column for the survival probabilities
surv_long$surv <- surv[as.numeric(surv_long$surv_id)]

# Add columns upper, lower, std.err, and strata contianing NA values
surv_long[, c("upper", "lower", "std.err", "strata")] <- NA

# Plot survival curves
ggsurvplot_df(surv_long, surv.geom = geom_line,
              linetype = "horTh", color = "dist", legend.title = NULL)
          
```

## The Cox model

### Intro
Semiparametric model (≠Weibull model, fully parametric). It means less strict distributional assumptions. 

Also called the proportional hazards assumption. Both the Cox and the Weibull models assume proportional harzards: the instantaneous rate of occurence of the event is assumed to be proportional (not seen in details here). In brief, the survival curves can't cross. If the curves do cross, none of the two models is appropriate. 

```{r}
cxmod <- coxph(Surv(time, cens) ~ horTh, data = GBSG2)
#summary(cxmod)
coef(cxmod) #no intercept ≠ wbmod
```
No intercept compared to the Weibull model, as it's semi parametric.
Contrary to Weibull model, a negative coefficients stands for a positive effect on duration time.  

### Visualization
Steps to produce visualization
- compute Cox model
- decide on "imaginary patients (combination of tsize and horTh)
- compute survival curves for imaginary patients
- create data.frame with survival curve information
- plot

```{r}
# Cox model
coxmd <- coxph(Surv(time, cens)~horTh + tsize, data = GBSG2)

# Imaginary patients
newdat <- expand.grid(
  horTh = levels(GBSG2$horTh),
  tsize = quantile(GBSG2$tsize, probs = c(.25, .5, .75)))
rownames(newdata)<-letters[1:6]  

# Inspect newdat
newdat
```

```{r}
# compute survival curves
cxsf <- survfit(coxmd, data = GBSG2, newdata = newdat, conf.type = "none")
head(cxsf$surv) 
# Look at the head() of the time points to see which time points head(cxsf$surv) corresponds to.
head(cxsf$time)
```
```{r}
#compute dataframe
surv_cxmod0 <- surv_summary(cxsf)
head(surv_cxmod0)

# get a character vector of patient letters
pid <- as.character(surv_cxmod0$strata)

# Multiple of the rows in newdat so that it fits with surv_cxmod0
m_newdat = newdat[pid,]

# Add patient info to data.frame
surv_cxmod0 <- cbind(surv_cxmod0, m_newdat)
head(surv_cxmod0)
```

```{r}
# Plot
ggsurvplot_df(surv_cxmod0, linetype="horTh", color="tsize",
              censor=FALSE, legend.title=NULL) #removing censor for readablity
```


### What we learned

Concepts
- Why survival methods 
- censoring
- survival curve

Methods
- Kaplan-Meier estimate
- Weibull model
- Cox model

Possible next steps:
- Mathematical details
- Interpretation of effect estimates
- Inference

Review Ex
Why "imaginary patients"? 
Because survival curves can only be plotted for given covariate combinations.


### Capstone
#### Cox Model
```{r}
# Remove the 1 row with a NA for our explanatory variable
#lung2=lung[!is.na(lung$ph.karno),]

# Compute Cox model and survival curves
cxmod <- coxph(Surv(time, status) ~ ph.karno, data = lung)

new_lung <- data.frame(ph.karno = c(60,70,80,90)) #imaginary patients
cxsf <- survfit(cxmod, data = lung, newdata = new_lung, conf.type = "none")


# Use the summary of cxsf to take a vector of patient IDs
surv_cxmod0 <- surv_summary(cxsf)
pid <- as.character(surv_cxmod0$strata)

# Duplicate rows in newdat to fit with surv_cxmod0 and add them in
m_newdata = new_lung[pid,, drop=FALSE]
surv_cxmod <- cbind(surv_cxmod0, m_newdata)
head(surv_cxmod)
```
```{r}
#plot
ggsurvplot_df(surv_cxmod, color="ph.karno", legend.title = NULL, censor=FALSE)
```

We can see that performance score do have an effect on survival probability. 

#### Comparing Survival Curves

Now, let's take a look at the survival curve of all individuals using the Kaplan-Meier estimate and compare it to the curve of a Cox model that takes performance into account.

```{r}
# Compute Kaplan-Meier curve
km <- survfit(Surv(time, status) ~ 1, data=lung)

# Compute Cox Model
coxmod <- coxph(Surv(time, status) ~ ph.karno, data=lung)

# Compute Cox model survival curves
new_lung <- data.frame(ph.karno = c(60,70,80,90))
cxsf <- survfit(coxmod, data = lung, newdata = new_lung, conf.type = "none")

# Plot Kaplan-Meier curve
ggsurvplot(km, conf.int = FALSE)

# Plot Cox model survival curves
## for Cox models, you can just enter the survfit() output into ggsurvplot() instead of
##creating the needed data frame yourself and plugging it into ggsurvplot_df().
ggsurvplot(cxsf, censor = FALSE)
```

